---
title: "MS Disability Progression - Weeks 1-5"
subtitle: "MSc Data Science | Student: 25044419"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)
```

---

# Setup: Load Libraries

```{r libraries}
library(tidyverse)
library(lubridate)
library(naniar)
library(skimr)
library(knitr)
library(kableExtra)
library(ggpubr)
library(scales)
library(viridis)
```

---

# Load the Data

```{r load-data}
patients   <- read_csv("patients.csv")
visits     <- read_csv("visits.csv")
relapses   <- read_csv("relapses.csv")
treatments <- read_csv("treatments.csv")

cat("patients   :", nrow(patients),   "rows,", ncol(patients),   "columns\n")
cat("visits     :", nrow(visits),     "rows,", ncol(visits),     "columns\n")
cat("relapses   :", nrow(relapses),   "rows,", ncol(relapses),   "columns\n")
cat("treatments :", nrow(treatments), "rows,", ncol(treatments), "columns\n")
```

---

# patients.csv - one row per patient

## Structure and column check

```{r patients-structure}
glimpse(patients)
```

## Select key columns

```{r patients-key}
patients_key <- patients %>%
  select(
    PATIENT_ID,
    GENDER,
    BIRTH_DATE,
    SYMPTOMS_DATE,
    MS_DIAGNOSIS_DATE,
    MSCOURSE_1,
    MSCOURSE_1_DATE,
    MSCOURSE_2,
    MSCOURSE_2_DATE,
    DEAD
  )

patients_key %>%
  head(10) %>%
  kable(caption = "patients.csv - first 10 rows (key columns only)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Convert dates

```{r patients-dates}
# dates are DD.MM.YYYY, use dmy() not as_date()
patients_key <- patients_key %>%
  mutate(
    BIRTH_DATE        = dmy(BIRTH_DATE),
    SYMPTOMS_DATE     = dmy(SYMPTOMS_DATE),
    MS_DIAGNOSIS_DATE = dmy(MS_DIAGNOSIS_DATE),
    MSCOURSE_1_DATE   = dmy(MSCOURSE_1_DATE),
    MSCOURSE_2_DATE   = dmy(MSCOURSE_2_DATE)
  )

cat("Birth dates:     ", format(range(patients_key$BIRTH_DATE,        na.rm = TRUE)), "\n")
cat("Symptom onset:   ", format(range(patients_key$SYMPTOMS_DATE,     na.rm = TRUE)), "\n")
cat("Diagnosis dates: ", format(range(patients_key$MS_DIAGNOSIS_DATE, na.rm = TRUE)), "\n")
```

## Derive age at onset and diagnostic delay

```{r patients-age}
patients_key <- patients_key %>%
  mutate(
    age_at_onset = as.numeric(
      difftime(SYMPTOMS_DATE, BIRTH_DATE, units = "days")
    ) / 365.25,

    diagnostic_delay_yrs = as.numeric(
      difftime(MS_DIAGNOSIS_DATE, SYMPTOMS_DATE, units = "days")
    ) / 365.25,

    # negative delays = data entry errors
    diagnostic_delay_yrs = if_else(diagnostic_delay_yrs < 0, NA_real_, diagnostic_delay_yrs)
  )

cat("Age at onset - summary:\n")
summary(patients_key$age_at_onset)
cat("\nDiagnostic delay (years) - summary:\n")
summary(patients_key$diagnostic_delay_yrs)
```

## Gender distribution

```{r patients-gender}
patients_key %>%
  count(GENDER) %>%
  mutate(pct = n / sum(n)) %>%
  kable(col.names = c("Gender", "N", "Proportion"),
        digits    = 3,
        caption   = "Gender distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Disease course at onset

```{r patients-course}
patients_key %>%
  count(MSCOURSE_1, sort = TRUE) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  kable(col.names = c("MS Course", "N", "Percent"),
        caption   = "Initial disease course (MSCOURSE_1)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Censoring status

```{r patients-dead}
patients_key %>%
  count(DEAD) %>%
  mutate(
    label = ifelse(DEAD == 1 | DEAD == TRUE, "Deceased", "Alive/Censored"),
    pct   = percent(n / sum(n), accuracy = 0.1)
  ) %>%
  kable(col.names = c("DEAD", "N", "Label", "Percent"),
        caption   = "Censoring status") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Patient-level plots

```{r patients-plots, fig.height=8}
p1 <- ggplot(patients_key, aes(x = age_at_onset)) +
  geom_histogram(bins = 30, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  labs(title = "Age at Symptom Onset", x = "Age (years)", y = "Number of patients") +
  theme_minimal(base_size = 13)

p2 <- ggplot(patients_key %>% filter(!is.na(GENDER)),
             aes(x = GENDER, fill = GENDER)) +
  geom_bar(colour = "white", alpha = 0.85) +
  scale_fill_manual(values = c("#2E4D7B", "#C85250", "#5A8A5A")) +
  labs(title = "Gender Distribution", x = "", y = "Number of patients") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p3 <- ggplot(patients_key %>% filter(!is.na(MSCOURSE_1)),
             aes(x = fct_infreq(MSCOURSE_1), fill = MSCOURSE_1)) +
  geom_bar(colour = "white", alpha = 0.85) +
  scale_fill_viridis_d() +
  labs(title = "Initial Disease Course", x = "", y = "Number of patients") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))

p4 <- ggplot(patients_key %>% filter(!is.na(diagnostic_delay_yrs),
                                      diagnostic_delay_yrs >= 0,
                                      diagnostic_delay_yrs < 30),
             aes(x = diagnostic_delay_yrs)) +
  geom_histogram(bins = 25, fill = "#5A8A5A", colour = "white", alpha = 0.85) +
  labs(title = "Diagnostic Delay (Symptom Onset to Diagnosis)",
       x = "Years", y = "Number of patients") +
  theme_minimal(base_size = 13)

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```

---

# File 2: visits.csv - One Row Per Clinical Visit

## Basic structure

```{r visits-structure}
cat("Dimensions:", nrow(visits), "rows x", ncol(visits), "columns\n")
cat("Columns:\n")
cat(names(visits), sep = "\n")
```

## Select key columns

```{r visits-key}
visits_key <- visits %>%
  select(
    PATIENT_ID,
    DATE_OF_VISIT,
    EDSS,
    DURATION_OF_MS_AT_VISIT,
    DURATION_OF_MS_AT_VISIT_ROUNDED,
    FIRST_MSCOURSE,
    MSCOURSE_AT_VISIT
  ) %>%
  mutate(DATE_OF_VISIT = dmy(DATE_OF_VISIT))

cat("Visits date range:              ", format(range(visits_key$DATE_OF_VISIT, na.rm = TRUE)), "\n")
cat("Unique patients with visits:    ", n_distinct(visits_key$PATIENT_ID), "\n")
cat("Total visits:                   ", nrow(visits_key), "\n")
cat("Average visits per patient:     ", round(nrow(visits_key) / n_distinct(visits_key$PATIENT_ID), 1), "\n")
```

## Visits per patient

```{r visits-per-patient}
visits_per_patient <- visits_key %>%
  count(PATIENT_ID, name = "n_visits")

cat("Visits per patient - summary:\n")
summary(visits_per_patient$n_visits)

ggplot(visits_per_patient, aes(x = n_visits)) +
  geom_histogram(bins = 30, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  geom_vline(aes(xintercept = median(n_visits)),
             colour = "#C85250", linetype = "dashed", linewidth = 1) +
  annotate("text",
           x     = median(visits_per_patient$n_visits) + 1,
           y     = Inf, vjust = 2,
           label = paste("Median =", median(visits_per_patient$n_visits)),
           colour = "#C85250", size = 4) +
  labs(title = "Distribution of Number of Visits per Patient",
       x = "Number of visits", y = "Number of patients") +
  theme_minimal(base_size = 13)
```

## Irregular visit spacing

```{r visit-spacing}
visit_gaps <- visits_key %>%
  arrange(PATIENT_ID, DATE_OF_VISIT) %>%
  group_by(PATIENT_ID) %>%
  mutate(
    prev_visit = lag(DATE_OF_VISIT),
    gap_days   = as.numeric(DATE_OF_VISIT - prev_visit),
    gap_months = gap_days / 30.44
  ) %>%
  ungroup() %>%
  filter(!is.na(gap_days))

cat("Inter-visit gap (days) - summary:\n")
summary(visit_gaps$gap_days)
cat("\nInter-visit gap (months) - summary:\n")
summary(visit_gaps$gap_months)

ggplot(visit_gaps %>% filter(gap_days < 730), aes(x = gap_months)) +
  geom_histogram(bins = 40, fill = "#5A8A5A", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 6,  colour = "#C85250", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 12, colour = "#2E4D7B", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 6.5,  y = Inf, vjust = 2,
           label = "6 months",  colour = "#C85250", size = 3.5) +
  annotate("text", x = 12.5, y = Inf, vjust = 2,
           label = "12 months", colour = "#2E4D7B", size = 3.5) +
  labs(title    = "Distribution of Inter-Visit Gaps (months)",
       subtitle = "Confirms irregular visit spacing - visits are NOT evenly scheduled",
       x = "Gap between consecutive visits (months)", y = "Count") +
  theme_minimal(base_size = 13)
```

## EDSS - primary outcome measure

```{r edss-summary}
cat("EDSS - summary:\n")
summary(visits_key$EDSS)
cat("EDSS missing:", sum(is.na(visits_key$EDSS)), "of", nrow(visits_key),
    "(", round(100 * mean(is.na(visits_key$EDSS)), 1), "%)\n")

# is missingness spread across patients or concentrated?
visits_key %>%
  mutate(edss_missing = is.na(EDSS)) %>%
  group_by(PATIENT_ID) %>%
  summarise(pct_missing = mean(edss_missing)) %>%
  summary()

visits_key %>%
  filter(!is.na(EDSS)) %>%
  ggplot(aes(x = EDSS)) +
  geom_histogram(binwidth = 0.5, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  scale_x_continuous(breaks = seq(0, 10, by = 1)) +
  labs(title    = "EDSS Distribution Across All Visits",
       subtitle = paste0("n = ", sum(!is.na(visits_key$EDSS)), " visits with EDSS recorded",
                         " | ", round(100 * mean(is.na(visits_key$EDSS)), 1), "% missing"),
       x = "EDSS Score", y = "Number of visits") +
  theme_minimal(base_size = 13) +
  annotate("text", x = 8, y = Inf, vjust = 2,
           label  = "Expected bimodal:\nclustering at low\nand moderate scores",
           size   = 3.5, colour = "#C85250")
```

## EDSS over time

```{r edss-over-time}
edss_trajectory <- visits_key %>%
  filter(!is.na(EDSS),
         !is.na(DURATION_OF_MS_AT_VISIT_ROUNDED),
         DURATION_OF_MS_AT_VISIT_ROUNDED <= 20) %>%
  group_by(DURATION_OF_MS_AT_VISIT_ROUNDED) %>%
  summarise(
    median_edss = median(EDSS),
    q25         = quantile(EDSS, 0.25),
    q75         = quantile(EDSS, 0.75),
    n_visits    = n()
  )

ggplot(edss_trajectory, aes(x = DURATION_OF_MS_AT_VISIT_ROUNDED)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#2E4D7B", alpha = 0.2) +
  geom_line(aes(y = median_edss), colour = "#2E4D7B", linewidth = 1.2) +
  geom_point(aes(y = median_edss, size = n_visits), colour = "#2E4D7B") +
  scale_size_continuous(range = c(2, 6), name = "N visits") +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), limits = c(0, 10)) +
  labs(title    = "Median EDSS by Years Since Symptom Onset",
       subtitle = "Shaded band = IQR (25th to 75th percentile)",
       x = "Years since symptom onset", y = "EDSS") +
  theme_minimal(base_size = 13)
```

## Missing data in visits.csv

```{r visits-missing}
missing_summary <- visits_key %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

missing_summary %>%
  kable(col.names = c("Column", "% Missing"),
        digits    = 1,
        caption   = "Missing data by column in visits.csv (key columns)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(missing_summary$pct_missing > 30), background = "#FCE4D6")
```

## EDSS missingness over disease duration

```{r edss-missing-over-time}
edss_missing_time <- visits_key %>%
  filter(!is.na(DURATION_OF_MS_AT_VISIT_ROUNDED),
         DURATION_OF_MS_AT_VISIT_ROUNDED <= 20) %>%
  group_by(DURATION_OF_MS_AT_VISIT_ROUNDED) %>%
  summarise(
    n_total     = n(),
    n_missing   = sum(is.na(EDSS)),
    pct_missing = 100 * n_missing / n_total
  )
print(edss_missing_time, n = 21)
as.data.frame(edss_missing_time)

ggplot(edss_missing_time,
       aes(x = DURATION_OF_MS_AT_VISIT_ROUNDED, y = pct_missing)) +
  geom_col(fill = "#C85250", colour = "white", alpha = 0.85) +
  geom_hline(yintercept = 45, linetype = "dashed",
             colour = "#2E4D7B", linewidth = 1) +
  annotate("text", x = 15, y = 47,
           label  = "Overall 45% missing",
           colour = "#2E4D7B", size = 3.5) +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +
  labs(title    = "EDSS Missingness by Year of Disease Duration",
       subtitle = "If missingness increases over time, this supports MNAR - sicker patients dropping out",
       x = "Years since symptom onset", y = "% of visits missing EDSS") +
  theme_minimal(base_size = 13)
```

---

# File 3: relapses.csv - One Row Per Relapse Event

## Basic structure

```{r relapses-structure}
glimpse(relapses)
```

## Counts and distribution

```{r relapses-key}
relapses_key <- relapses %>%
  select(PATIENT_ID, DATE_OF_ONSET) %>%
  mutate(DATE_OF_ONSET = dmy(DATE_OF_ONSET))

relapses_per_patient <- relapses_key %>%
  count(PATIENT_ID, name = "n_relapses")

cat("Total relapses:                     ", nrow(relapses_key), "\n")
cat("Patients with at least one relapse: ", n_distinct(relapses_key$PATIENT_ID), "\n")
cat("Patients with NO recorded relapses: ",
    n_distinct(patients$PATIENT_ID) - n_distinct(relapses_key$PATIENT_ID), "\n")
cat("\nRelapses per patient - summary:\n")
summary(relapses_per_patient$n_relapses)
```

## Relapse plots

```{r relapses-plot}
p1 <- ggplot(relapses_per_patient, aes(x = n_relapses)) +
  geom_histogram(bins = 20, fill = "#C85250", colour = "white", alpha = 0.85) +
  labs(title = "Relapses per Patient",
       x = "Number of relapses", y = "Number of patients") +
  theme_minimal(base_size = 13)

relapses_with_time <- relapses_key %>%
  left_join(patients_key %>% select(PATIENT_ID, SYMPTOMS_DATE),
            by = "PATIENT_ID") %>%
  mutate(
    time_since_onset_yrs = as.numeric(
      difftime(DATE_OF_ONSET, SYMPTOMS_DATE, units = "days")
    ) / 365.25
  ) %>%
  filter(!is.na(time_since_onset_yrs),
         time_since_onset_yrs >= 0,
         time_since_onset_yrs <= 20)

p2 <- ggplot(relapses_with_time, aes(x = time_since_onset_yrs)) +
  geom_histogram(bins = 30, fill = "#C85250", colour = "white", alpha = 0.85) +
  labs(title = "Relapse Timing Since Symptom Onset",
       x = "Years since onset", y = "Number of relapses") +
  theme_minimal(base_size = 13)

ggarrange(p1, p2, ncol = 2)
```

---

# File 4: treatments.csv - One Row Per Treatment Episode

## Basic structure

```{r treatments-structure}
glimpse(treatments)
```

## DMT types

```{r treatments-dmt}
treatments_key <- treatments %>%
  select(PATIENT_ID, DMT, START_DATE, END_DATE) %>%
  mutate(
    START_DATE = dmy(START_DATE),
    END_DATE   = dmy(END_DATE)
  )

treatments_key %>%
  count(DMT, sort = TRUE) %>%
  head(20) %>%
  kable(col.names = c("DMT", "N episodes"),
        caption   = "Top 20 most common DMTs") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Treatment episodes per patient

```{r treatments-episodes}
episodes_per_patient <- treatments_key %>%
  count(PATIENT_ID, name = "n_episodes")

cat("DMT episodes per patient - summary:\n")
summary(episodes_per_patient$n_episodes)
cat("\nPatients with no treatment recorded:",
    n_distinct(patients$PATIENT_ID) - n_distinct(treatments_key$PATIENT_ID), "\n")
```

---

# Follow-Up Analysis

```{r followup}
followup <- visits_key %>%
  group_by(PATIENT_ID) %>%
  summarise(
    first_visit = min(DATE_OF_VISIT, na.rm = TRUE),
    last_visit  = max(DATE_OF_VISIT, na.rm = TRUE),
    n_visits    = n()
  ) %>%
  left_join(patients_key %>% select(PATIENT_ID, SYMPTOMS_DATE),
            by = "PATIENT_ID") %>%
  mutate(
    followup_from_onset_yrs  = as.numeric(
      difftime(last_visit, SYMPTOMS_DATE, units = "days")
    ) / 365.25,
    followup_in_registry_yrs = as.numeric(
      difftime(last_visit, first_visit, units = "days")
    ) / 365.25
  )

cat("Follow-up from onset to last visit (years):\n")
summary(followup$followup_from_onset_yrs)

cat("\nPatients with 2+ years of follow-up:", sum(followup$followup_from_onset_yrs >= 2, na.rm = TRUE), "\n")
cat("Patients with 3+ years of follow-up:", sum(followup$followup_from_onset_yrs >= 3, na.rm = TRUE), "\n")
cat("Patients with 5+ years of follow-up:", sum(followup$followup_from_onset_yrs >= 5, na.rm = TRUE), "\n")
cat("Patients with 7+ years of follow-up:", sum(followup$followup_from_onset_yrs >= 7, na.rm = TRUE), "\n")
cat("Patients with 20+ years of follow-up:", sum(followup$followup_from_onset_yrs >= 20, na.rm = TRUE), "\n")

ggplot(followup %>% filter(!is.na(followup_from_onset_yrs)),
       aes(x = followup_from_onset_yrs)) +
  geom_histogram(bins = 30, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 2, colour = "#C85250", linetype = "dashed", linewidth = 1.2) +
  geom_vline(xintercept = 5, colour = "#5A8A5A", linetype = "dashed", linewidth = 1.2) +
  geom_vline(xintercept = 7, colour = "#E8A020", linetype = "dashed", linewidth = 1.2) +
  annotate("text", x = 2.2, y = Inf, vjust = 1, label = "2 yrs", colour = "#C85250", size = 3.5) +
  annotate("text", x = 5.2, y = Inf, vjust = 1, label = "5 yrs", colour = "#5A8A5A", size = 3.5) +
  annotate("text", x = 7.2, y = Inf, vjust = 1, label = "7 yrs", colour = "#E8A020", size = 3.5) +
  labs(title    = "Distribution of Patient Follow-Up Length (from symptom onset)",
       subtitle = "Dashed lines show proposed feature and prediction windows",
       x = "Years from symptom onset to last visit", y = "Number of patients") +
  theme_minimal(base_size = 13)
```

## Patients with EDSS data in year 0–2

```{r edss-in-window}
patients_with_early_edss <- visits_key %>%
  filter(!is.na(EDSS),
         DURATION_OF_MS_AT_VISIT >= 0,
         DURATION_OF_MS_AT_VISIT <= 2) %>%
  distinct(PATIENT_ID)

cat("Patients with at least 1 EDSS in year 0-2:",
    nrow(patients_with_early_edss), "of", n_distinct(visits_key$PATIENT_ID), "\n")

patients_with_2_edss <- visits_key %>%
  filter(!is.na(EDSS),
         DURATION_OF_MS_AT_VISIT >= 0,
         DURATION_OF_MS_AT_VISIT <= 2) %>%
  count(PATIENT_ID) %>%
  filter(n >= 2)

cat("Patients with 2+ EDSS in year 0-2:", nrow(patients_with_2_edss), "\n")
```

---

# Missing Data Overview - All Files

```{r missing-overview, fig.height=10}
analysis_vars <- visits_key %>%
  left_join(patients_key %>% select(PATIENT_ID, GENDER, age_at_onset,
                                     MSCOURSE_1, DEAD),
            by = "PATIENT_ID")

gg_miss_var(analysis_vars %>% select(-PATIENT_ID), show_pct = TRUE) +
  labs(title    = "Missing Data - Combined Key Variables",
       subtitle = "Percentage of values missing per variable") +
  theme_minimal(base_size = 12)
```

---

# Summary - Key Findings (End of Week 1)

```{r summary-table}
summary_findings <- tibble(
  Category = c(
    "Total patients",
    "Total clinical visits",
    "Average visits per patient",
    "Follow-up range (years from onset)",
    "Patients with 2+ years follow-up",
    "Patients with 5+ years follow-up",
    "EDSS missing (all visits)",
    "Patients with 2+ EDSS values in year 0-2",
    "Total relapses recorded",
    "Total DMT episodes recorded"
  ),
  Value = c(
    n_distinct(patients$PATIENT_ID),
    nrow(visits),
    round(nrow(visits) / n_distinct(visits$PATIENT_ID), 1),
    paste(round(min(followup$followup_from_onset_yrs, na.rm = TRUE), 1), "to",
          round(max(followup$followup_from_onset_yrs, na.rm = TRUE), 1)),
    sum(followup$followup_from_onset_yrs >= 2, na.rm = TRUE),
    sum(followup$followup_from_onset_yrs >= 5, na.rm = TRUE),
    paste0(round(100 * mean(is.na(visits$EDSS)), 1), "%"),
    nrow(patients_with_2_edss),
    nrow(relapses),
    nrow(treatments)
  )
)

summary_findings %>%
  kable(caption = "Key Dataset Statistics - End of Week 1") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Week 2 - Cleaning and Outcome Definition

## DMT harmonisation

```{r dmt-cleaning}
# TREATMENT column has the drug name even where DMT is NA
# VISIT_ID == "SYMPTOMATIC" = relapse/symptom treatment, not a DMT
treatments_key <- treatments %>%
  select(PATIENT_ID, DMT, TREATMENT, VISIT_ID, START_DATE, END_DATE) %>%
  mutate(
    START_DATE = dmy(START_DATE),
    END_DATE   = dmy(END_DATE),

    DMT_clean = case_when(
      VISIT_ID == "SYMPTOMATIC"          ~ "Symptomatic",

      # same drug, different dose - merge on name
      str_detect(DMT, "OCREVUS")         ~ "OCREVUS",
      str_detect(DMT, "TYSABRI")         ~ "TYSABRI",
      str_detect(DMT, "GILENYA")         ~ "GILENYA",
      str_detect(DMT, "TECFIDERA")       ~ "TECFIDERA",
      str_detect(DMT, "COPAXONE")        ~ "COPAXONE",
      str_detect(DMT, "AVONEX")          ~ "AVONEX",
      str_detect(DMT, "BETAFERON")       ~ "BETAFERON",

      # same drug, different name
      # NATALIZUMAB = TYSABRI (checked EMA product page)
      DMT == "NATALIZUMAB"               ~ "TYSABRI",

      !is.na(DMT)                        ~ DMT,
      TRUE                               ~ "Other"
    )
  )

treatments_key %>%
  count(DMT_clean, sort = TRUE) %>%
  kable(col.names = c("DMT", "N episodes"),
        caption   = "DMT episodes after harmonisation") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Course type grouping

```{r course-grouping}
# CIS = 307, PP = 33, NA = 60
# No RR in this dataset, CIS is the pre-confirmation label (converts to RR)
# PP = 33, above threshold of 10, kept separate
# NA = 60 - relabelled as "MS" (diagnosed but no course recorded at onset)

patients_key %>%
  filter(is.na(MSCOURSE_1)) %>%
  select(PATIENT_ID, MSCOURSE_1, MSCOURSE_2, MS_DIAGNOSIS_DATE) %>%
  head(10) %>%
  kable(caption = "Patients with no MS course recorded") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Feature window - compare 2 vs 3 year window

```{r window-recheck}
# widen to 0-3 years if 0-2 gives only 74 patients
# prediction window shifts to years 4-6 accordingly
patients_with_2_edss_3yr <- visits_key %>%
  filter(!is.na(EDSS),
         DURATION_OF_MS_AT_VISIT >= 0,
         DURATION_OF_MS_AT_VISIT <= 3) %>%
  count(PATIENT_ID) %>%
  filter(n >= 2)

cat("Eligible patients - 2yr window (Week 1 finding):", length(patients_with_2_edss$PATIENT_ID), "\n")
cat("Eligible patients - 3yr window:                 ", nrow(patients_with_2_edss_3yr), "\n")
cat("Gain from widening window:                      ",
    nrow(patients_with_2_edss_3yr) - nrow(patients_with_2_edss), "patients\n")
```

## Baseline EDSS per patient

```{r baseline-edss}
# baseline = first EDSS value recorded from symptom onset
# CDP threshold depends on baseline (Kappos et al. 2018, Neurology)
baseline_edss <- visits_key %>%
  filter(!is.na(EDSS), DURATION_OF_MS_AT_VISIT >= 0) %>%
  arrange(PATIENT_ID, DATE_OF_VISIT) %>%
  group_by(PATIENT_ID) %>%
  slice(1) %>%
  ungroup() %>%
  select(PATIENT_ID, baseline_edss = EDSS) %>%
  mutate(cdp_threshold = if_else(baseline_edss >= 5.5, 0.5, 1.0))

cat("Baseline EDSS - summary:\n")
summary(baseline_edss$baseline_edss)
cat("\nPatients with threshold 0.5 (baseline >= 5.5):",
    sum(baseline_edss$cdp_threshold == 0.5), "\n")
cat("Patients with threshold 1.0 (baseline < 5.5): ",
    sum(baseline_edss$cdp_threshold == 1.0), "\n")
```

## Define CDP outcome (years 4–6)

```{r cdp-outcome}
# step 1: visits in prediction window with EDSS
prediction_visits <- visits_key %>%
  filter(DURATION_OF_MS_AT_VISIT >= 4,
         DURATION_OF_MS_AT_VISIT <= 6,
         !is.na(EDSS)) %>%
  left_join(baseline_edss, by = "PATIENT_ID") %>%
  mutate(
    edss_increase   = EDSS - baseline_edss,
    meets_threshold = edss_increase >= cdp_threshold
  )

# step 2: unconfirmed CDP - at least one visit meets threshold
cdp_unconfirmed <- prediction_visits %>%
  group_by(PATIENT_ID) %>%
  summarise(cdp = any(meets_threshold, na.rm = TRUE)) %>%
  ungroup()

cat("Patients with visits in prediction window: ", nrow(cdp_unconfirmed), "\n")
cat("CDP events (unconfirmed):                  ", sum(cdp_unconfirmed$cdp), "\n")
cat("CDP rate   (unconfirmed):                  ",
    round(100 * mean(cdp_unconfirmed$cdp), 1), "%\n")

# step 3: confirmed CDP - increase at two consecutive visits
cdp_confirmed <- prediction_visits %>%
  arrange(PATIENT_ID, DATE_OF_VISIT) %>%
  group_by(PATIENT_ID) %>%
  mutate(confirmed = meets_threshold & lag(meets_threshold, default = FALSE)) %>%
  summarise(cdp = any(confirmed, na.rm = TRUE)) %>%
  ungroup()

cat("\nCDP events (confirmed at 2 consecutive): ", sum(cdp_confirmed$cdp), "\n")
cat("CDP rate   (confirmed):                   ",
    round(100 * mean(cdp_confirmed$cdp), 1), "%\n")

# Decision: use unconfirmed - confirmed gives only 13 events (too few)
# unconfirmed gives 30 events, sufficient for modelling
# note: cdp_confirmed is not used anywhere past this point - it exists
# only to justify the decision above
```

## Handle deceased patients

```{r censor-deaths}
last_visits <- visits_key %>%
  group_by(PATIENT_ID) %>%
  summarise(last_visit = max(DATE_OF_VISIT, na.rm = TRUE))

deceased_patients <- patients_key %>%
  filter(DEAD == TRUE) %>%
  left_join(last_visits, by = "PATIENT_ID") %>%
  mutate(followup_at_death = as.numeric(
    difftime(last_visit, SYMPTOMS_DATE, units = "days")
  ) / 365.25)

cat("Total deceased patients:                                ", nrow(deceased_patients), "\n")
cat("Deceased with follow-up under 6 years (within window): ",
    sum(deceased_patients$followup_at_death <= 6, na.rm = TRUE), "\n")
```

## Final modelling cohort

```{r modelling-cohort}
# inclusion criteria: 2+ EDSS values in years 0-3 (feature window), at least
# one visit with EDSS in years 4-6 (prediction window), not deceased within
# the prediction window

eligible_feature <- visits_key %>%
  filter(!is.na(EDSS),
         DURATION_OF_MS_AT_VISIT >= 0,
         DURATION_OF_MS_AT_VISIT <= 3) %>%
  count(PATIENT_ID) %>%
  filter(n >= 2) %>%
  pull(PATIENT_ID)

eligible_prediction <- visits_key %>%
  filter(DURATION_OF_MS_AT_VISIT >= 4,
         DURATION_OF_MS_AT_VISIT <= 6,
         !is.na(EDSS)) %>%
  distinct(PATIENT_ID) %>%
  pull(PATIENT_ID)

deceased_early <- deceased_patients %>%
  filter(followup_at_death <= 6) %>%
  pull(PATIENT_ID)

final_cohort <- intersect(eligible_feature, eligible_prediction)
final_cohort <- setdiff(final_cohort, deceased_early)

cat("Patients with 2+ EDSS in feature window:          ", length(eligible_feature), "\n")
cat("Patients with visits in prediction window:        ", length(eligible_prediction), "\n")
cat("Removed - deceased within prediction window:      ", length(deceased_early), "\n")
cat("Final modelling cohort:                           ", length(final_cohort), "\n")
```

## Included vs Excluded Patient Comparison

```{r inc-vs-exc-comparison}
excluded_cohort <- setdiff(patients_key$PATIENT_ID, final_cohort)

comparison_tbl <- patients_key %>%
  mutate(group = ifelse(PATIENT_ID %in% final_cohort, "Included", "Excluded")) %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean_age = round(mean(age_at_onset, na.rm = TRUE), 1),
    pct_female = round(100 * mean(GENDER == "F", na.rm = TRUE), 1),
    pct_cis = round(100 * mean(MSCOURSE_1 == "CIS", na.rm = TRUE), 1),
    pct_pp = round(100 * mean(MSCOURSE_1 == "PP", na.rm = TRUE), 1)
  )

comparison_tbl
```

## Follow-up Length by Inclusion Status

```{r followup-comparison}
followup_comparison <- visits_key %>%
  group_by(PATIENT_ID) %>%
  summarise(n_visits = n()) %>%
  mutate(group = ifelse(PATIENT_ID %in% final_cohort, "Included", "Excluded")) %>%
  group_by(group) %>%
  summarise(
    mean_visits = round(mean(n_visits, na.rm = TRUE), 1),
    median_visits = median(n_visits, na.rm = TRUE)
  )

followup_comparison
```
## CDP outcome within final cohort

```{r cdp-final}
cdp_final <- cdp_unconfirmed %>%
  filter(PATIENT_ID %in% final_cohort)

cat("Final cohort size:                  ", length(final_cohort), "\n")
cat("CDP events in final cohort:         ", sum(cdp_final$cdp), "\n")
cat("CDP rate in final cohort:           ", round(100 * mean(cdp_final$cdp), 1), "%\n")
cat("Non-CDP in final cohort:            ", sum(!cdp_final$cdp), "\n")

patients_no_outcome <- final_cohort[!final_cohort %in% cdp_final$PATIENT_ID]
cat("Patients with no assignable outcome:", length(patients_no_outcome), "\n")
```

## Alternative window - 0–4yr feature, 5–8yr prediction

> Exploratory check only. Not part of the final pipeline. Reported in weekly update.

```{r window-alternative}
eligible_feature_4yr <- visits_key %>%
  filter(!is.na(EDSS),
         DURATION_OF_MS_AT_VISIT >= 0,
         DURATION_OF_MS_AT_VISIT <= 4) %>%
  count(PATIENT_ID) %>%
  filter(n >= 2) %>%
  pull(PATIENT_ID)

eligible_prediction_58 <- visits_key %>%
  filter(DURATION_OF_MS_AT_VISIT >= 5,
         DURATION_OF_MS_AT_VISIT <= 8) %>%
  distinct(PATIENT_ID) %>%
  pull(PATIENT_ID)

cohort_4yr <- intersect(eligible_feature_4yr, eligible_prediction_58)

prediction_visits_58 <- visits_key %>%
  filter(DURATION_OF_MS_AT_VISIT >= 5,
         DURATION_OF_MS_AT_VISIT <= 8,
         !is.na(EDSS)) %>%
  left_join(baseline_edss, by = "PATIENT_ID") %>%
  mutate(
    edss_increase   = EDSS - baseline_edss,
    meets_threshold = edss_increase >= cdp_threshold
  )

cdp_58 <- prediction_visits_58 %>%
  group_by(PATIENT_ID) %>%
  summarise(cdp = any(meets_threshold, na.rm = TRUE)) %>%
  filter(PATIENT_ID %in% cohort_4yr)

cat("Feature 0-4yr, prediction 5-8yr:\n")
cat("Cohort size: ", length(cohort_4yr), "\n")
cat("CDP events:  ", sum(cdp_58$cdp), "\n")
cat("CDP rate:    ", round(100 * mean(cdp_58$cdp), 1), "%\n")
```

---

# Week 3 - Feature Engineering

> Feature window: years 0–3 from symptom onset. Prediction window: years 4–6.  
> Ordinal matrix is complete - no imputation needed.

## Finalise MS course grouping

```{r course-fix}
# Rename Unknown → "MS" (patients without a recorded disease course
# at onset who were nonetheless diagnosed)
patients_key <- patients_key %>%
  mutate(MSCOURSE_grouped = case_when(
    MSCOURSE_1 == "CIS" ~ "CIS",
    MSCOURSE_1 == "PP"  ~ "PP",
    is.na(MSCOURSE_1)   ~ "MS",
    TRUE                ~ "Other"
  ))

patients_key %>%
  count(MSCOURSE_grouped, sort = TRUE) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  kable(col.names = c("MS Course", "N", "Percent"),
        caption   = "MS course grouping (full cohort, pre-merge)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Feature window visits

```{r feature-visits}
feature_visits <- visits_key %>%
  filter(PATIENT_ID %in% final_cohort,
         !is.na(EDSS),
         DURATION_OF_MS_AT_VISIT >= 0,
         DURATION_OF_MS_AT_VISIT <= 3)

edss_counts <- feature_visits %>%
  count(PATIENT_ID, name = "n_edss_in_window")

cat("EDSS observations per patient in feature window:\n")
summary(edss_counts$n_edss_in_window)

patients_slope_eligible <- edss_counts %>%
  filter(n_edss_in_window >= 3) %>%
  pull(PATIENT_ID)

cat("\nPatients with 3+ EDSS values (slope computable): ",
    length(patients_slope_eligible), "\n")
cat("Patients with exactly 2 EDSS values (slope NA):  ",
    sum(edss_counts$n_edss_in_window == 2), "\n")
```

## Continuous EDSS features

```{r features-continuous}
edss_slope <- feature_visits %>%
  filter(PATIENT_ID %in% patients_slope_eligible) %>%
  group_by(PATIENT_ID) %>%
  summarise(
    edss_slope = coef(lm(EDSS ~ DURATION_OF_MS_AT_VISIT))[2]
  ) %>%
  ungroup()

edss_variability <- feature_visits %>%
  group_by(PATIENT_ID) %>%
  summarise(edss_sd = sd(EDSS, na.rm = TRUE)) %>%
  ungroup()

cat("EDSS slope - summary:\n")
summary(edss_slope$edss_slope)
cat("\nEDSS variability (SD) - summary:\n")
summary(edss_variability$edss_sd)
```

## Ordinal EDSS features

```{r features-ordinal}
ordinal_features <- feature_visits %>%
  arrange(PATIENT_ID, DATE_OF_VISIT) %>%
  group_by(PATIENT_ID) %>%
  summarise(
    n_worsening_steps = sum(diff(EDSS) >= 0.5, na.rm = TRUE),
    ordinal_change    = last(EDSS) - first(EDSS),
    max_edss          = max(EDSS, na.rm = TRUE)
  ) %>%
  ungroup()

cat("Worsening steps - summary:\n")
summary(ordinal_features$n_worsening_steps)
cat("\nOrdinal change (first to last) - summary:\n")
summary(ordinal_features$ordinal_change)
```

## Visit frequency in feature window

```{r feature-visit-freq}
visit_frequency <- visits_key %>%
  filter(PATIENT_ID %in% final_cohort,
         DURATION_OF_MS_AT_VISIT >= 0,
         DURATION_OF_MS_AT_VISIT <= 3) %>%
  group_by(PATIENT_ID) %>%
  summarise(
    total_visits  = n(),
    visit_freq_yr = n() / 3
  ) %>%
  ungroup()

cat("Visit frequency (per year) - summary:\n")
summary(visit_frequency$visit_freq_yr)
```

## Apply 30-day rule to relapses

```{r relapse-30day-rule}
# apply 30-day rule to MS-0122
# relapses within 30 days of the previous one count as the same episode
relapses_key_clean <- relapses_key %>%
  arrange(PATIENT_ID, DATE_OF_ONSET) %>%
  group_by(PATIENT_ID) %>%
  mutate(
    days_since_prev = as.numeric(DATE_OF_ONSET - lag(DATE_OF_ONSET)),
    keep            = is.na(days_since_prev) | days_since_prev >= 30
  ) %>%
  filter(keep) %>%
  ungroup() %>%
  select(PATIENT_ID, DATE_OF_ONSET)

# verify MS-0122 corrected count
ms0122_raw     <- relapses_key       %>% filter(PATIENT_ID == "MS-0122") %>% nrow()
ms0122_cleaned <- relapses_key_clean %>% filter(PATIENT_ID == "MS-0122") %>% nrow()

cat("MS-0122 relapses - raw:     ", ms0122_raw,     "\n")
cat("MS-0122 relapses - cleaned: ", ms0122_cleaned, "\n")
```

## Relapse count in feature window

```{r feature-relapses}
relapse_counts <- relapses_key_clean %>%
  left_join(patients_key %>% select(PATIENT_ID, SYMPTOMS_DATE),
            by = "PATIENT_ID") %>%
  mutate(
    time_since_onset = as.numeric(
      difftime(DATE_OF_ONSET, SYMPTOMS_DATE, units = "days")
    ) / 365.25
  ) %>%
  filter(PATIENT_ID %in% final_cohort,
         time_since_onset >= 0,
         time_since_onset <= 3) %>%
  count(PATIENT_ID, name = "relapse_count")

# patients with no relapses get 0, not NA
relapse_counts_full <- tibble(PATIENT_ID = final_cohort) %>%
  left_join(relapse_counts, by = "PATIENT_ID") %>%
  mutate(relapse_count = replace_na(relapse_count, 0))

cat("Relapse count in feature window - summary:\n")
summary(relapse_counts_full$relapse_count)
cat("Patients with zero relapses:", sum(relapse_counts_full$relapse_count == 0), "\n")

relapse_counts_full %>%
  arrange(desc(relapse_count)) %>%
  head(5) %>%
  kable(caption = "Top 5 patients by relapse count") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Merge small MS course groups

```{r course-merge}
# In the 74-patient modelling cohort: CIS = 66, MS = 4, PP = 4
# MS and PP both under threshold of 10 - merge into "Other"
patients_key <- patients_key %>%
  mutate(MSCOURSE_grouped = fct_collapse(
    factor(MSCOURSE_grouped),
    Other = c("MS", "PP")
  ))

patients_key %>%
  count(MSCOURSE_grouped, sort = TRUE) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  kable(col.names = c("MS Course", "N", "Percent"),
        caption   = "Final MS course grouping (full cohort, post-merge)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

## Assemble feature matrices

```{r feature-matrix-continuous}
features_continuous <- tibble(PATIENT_ID = final_cohort) %>%
  left_join(baseline_edss %>% select(PATIENT_ID, baseline_edss), by = "PATIENT_ID") %>%
  left_join(edss_slope,                                           by = "PATIENT_ID") %>%
  left_join(edss_variability,                                     by = "PATIENT_ID") %>%
  left_join(visit_frequency %>% select(PATIENT_ID, visit_freq_yr), by = "PATIENT_ID") %>%
  left_join(relapse_counts_full,                                  by = "PATIENT_ID") %>%
  left_join(patients_key %>% select(PATIENT_ID, age_at_onset,
                                     GENDER, MSCOURSE_grouped),  by = "PATIENT_ID") %>%
  left_join(cdp_final,                                            by = "PATIENT_ID") %>%
  mutate(
    GENDER           = factor(GENDER),
    MSCOURSE_grouped = factor(MSCOURSE_grouped),
    cdp              = as.integer(cdp)
  )

cat("Continuous feature matrix:", nrow(features_continuous), "rows x",
    ncol(features_continuous), "columns\n")
features_continuous %>%
  count(MSCOURSE_grouped) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  kable(caption = "MS course distribution in 74-patient modelling cohort") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r feature-matrix-ordinal}
features_ordinal <- tibble(PATIENT_ID = final_cohort) %>%
  left_join(baseline_edss %>% select(PATIENT_ID, baseline_edss), by = "PATIENT_ID") %>%
  left_join(ordinal_features,                                     by = "PATIENT_ID") %>%
  left_join(visit_frequency %>% select(PATIENT_ID, visit_freq_yr), by = "PATIENT_ID") %>%
  left_join(relapse_counts_full,                                  by = "PATIENT_ID") %>%
  left_join(patients_key %>% select(PATIENT_ID, age_at_onset,
                                     GENDER, MSCOURSE_grouped),  by = "PATIENT_ID") %>%
  left_join(cdp_final,                                            by = "PATIENT_ID") %>%
  mutate(
    GENDER           = factor(GENDER),
    MSCOURSE_grouped = factor(MSCOURSE_grouped),
    cdp              = as.integer(cdp)
  )

cat("Ordinal feature matrix:", nrow(features_ordinal), "rows x",
    ncol(features_ordinal), "columns\n")
```

## Missing values per feature

```{r feature-missing}
# continuous matrix
features_continuous %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  mutate(pct_missing = round(100 * n_missing / nrow(features_continuous), 1)) %>%
  arrange(desc(n_missing)) %>%
  kable(caption = "Missing values - continuous matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# ordinal matrix
features_ordinal %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  mutate(pct_missing = round(100 * n_missing / nrow(features_ordinal), 1)) %>%
  arrange(desc(n_missing)) %>%
  kable(caption = "Missing values - ordinal matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Feature summaries

```{r feature-summary}
cat("=== Continuous feature matrix ===\n")
features_continuous %>% select(-PATIENT_ID, -cdp) %>% summary()

cat("\n=== Ordinal feature matrix ===\n")
features_ordinal %>% select(-PATIENT_ID, -cdp) %>% summary()
```

## Feature distributions

```{r feature-plots, fig.height=10}
p1 <- ggplot(features_continuous %>% filter(!is.na(baseline_edss)),
             aes(x = baseline_edss)) +
  geom_histogram(bins = 15, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  labs(title = "Baseline EDSS", x = "EDSS", y = "Patients") +
  theme_minimal(base_size = 12)

p2 <- ggplot(features_continuous %>% filter(!is.na(edss_slope)),
             aes(x = edss_slope)) +
  geom_histogram(bins = 15, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, colour = "#C85250", linetype = "dashed") +
  labs(title = "EDSS Slope", x = "EDSS units per year", y = "Patients") +
  theme_minimal(base_size = 12)

p3 <- ggplot(features_continuous %>% filter(!is.na(edss_sd)),
             aes(x = edss_sd)) +
  geom_histogram(bins = 15, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  labs(title = "EDSS Variability (SD)", x = "SD", y = "Patients") +
  theme_minimal(base_size = 12)

p4 <- ggplot(features_continuous %>% filter(!is.na(age_at_onset)),
             aes(x = age_at_onset)) +
  geom_histogram(bins = 20, fill = "#5A8A5A", colour = "white", alpha = 0.85) +
  labs(title = "Age at Onset", x = "Years", y = "Patients") +
  theme_minimal(base_size = 12)

p5 <- ggplot(features_continuous, aes(x = relapse_count)) +
  geom_histogram(bins = 10, fill = "#C85250", colour = "white", alpha = 0.85) +
  labs(title = "Relapse Count in Feature Window",
       x = "Number of relapses", y = "Patients") +
  theme_minimal(base_size = 12)

p6 <- ggplot(features_continuous %>% filter(!is.na(MSCOURSE_grouped)),
             aes(x = MSCOURSE_grouped, fill = MSCOURSE_grouped)) +
  geom_bar(colour = "white", alpha = 0.85) +
  scale_fill_manual(values = c("#2E4D7B", "#C85250", "#5A8A5A")) +
  labs(title = "MS Course", x = "", y = "Patients") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

ggarrange(p1, p2, p3, p4, p5, p6, ncol = 2, nrow = 3)
```

## CDP balance check

```{r cdp-balance}
cat("Continuous matrix - outcome distribution:\n")
features_continuous %>%
  count(cdp) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  kable(col.names = c("CDP", "N", "Percent"),
        caption   = "Outcome balance - continuous matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

cat("\nOrdinal matrix - outcome distribution:\n")
features_ordinal %>%
  count(cdp) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  kable(col.names = c("CDP", "N", "Percent"),
        caption   = "Outcome balance - ordinal matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Save final matrices

```{r save-matrices}

saveRDS(features_continuous, "features_continuous.rds")
saveRDS(features_ordinal,    "features_ordinal.rds")

cat("Matrices saved.\n")
cat("Continuous:", nrow(features_continuous), "patients,",
    ncol(features_continuous) - 2, "features\n")
cat("Ordinal:   ", nrow(features_ordinal), "patients,",
    ncol(features_ordinal) - 2, "features\n")
```
---

# Week 4 - Imputation Pipeline

> Continuous matrix has `r sum(is.na(features_continuous$edss_slope))` missing edss_slope values (`r round(100 * mean(is.na(features_continuous$edss_slope)), 1)`%).
> Four strategies applied: complete case, median imputation, MICE, joint model (jomo).
> Ordinal matrix is complete - no imputation needed.

## Load libraries for imputation

```{r imputation-libraries}
library(mice)
library(jomo)
```

## Strategy 1 - Complete case analysis

```{r impute-cc}
# drop patients with any missing feature
# only edss_slope is missing so this removes the 22 patients with 2 EDSS values
features_cc <- features_continuous %>%
  filter(complete.cases(select(., -PATIENT_ID)))

cat("Complete case cohort:    ", nrow(features_cc), "patients\n")
cat("CDP events retained:     ", sum(features_cc$cdp), "\n")
cat("Patients removed:        ", nrow(features_continuous) - nrow(features_cc), "\n")
cat("CDP rate after CC:       ", round(100 * mean(features_cc$cdp), 1), "%\n")
```

## Strategy 2 - Median imputation

```{r impute-median}
# replace missing edss_slope with column median
median_slope <- median(features_continuous$edss_slope, na.rm = TRUE)

features_median <- features_continuous %>%
  mutate(edss_slope = if_else(is.na(edss_slope), median_slope, edss_slope))

cat("Median imputed cohort:     ", nrow(features_median), "patients\n")
cat("Slopes imputed:          ", sum(is.na(features_continuous$edss_slope)), "\n")
cat("Median used for imputation:", round(median_slope, 4), "\n")

# verify no missing values remain
cat("Missing after imputation:", sum(is.na(features_median$edss_slope)), "\n")
```

## Strategy 3 - MICE

```{r impute-mice}
# edss_slope is the only missing variable
# edss_slope is a continuous derived feature (regression coefficient)
# pmm is correct for continuous variables
# polr (proportional odds logistic regression) would apply if imputing
# raw EDSS scores directly - relevant for ordinal arm if needed later

mice_data <- features_continuous %>%
  select(-PATIENT_ID)

mice_methods <- make.method(mice_data)
mice_methods["edss_slope"] <- "pmm"

pred_matrix <- make.predictorMatrix(mice_data)

set.seed(42)
mice_imp <- mice(
  mice_data,
  m               = 20,
  method          = mice_methods,
  predictorMatrix = pred_matrix,
  printFlag       = FALSE
)

cat("MICE completed\n")
cat("Number of imputed datasets:", mice_imp$m, "\n")
cat("Variables imputed:", paste(names(mice_imp$method[mice_imp$method != ""]), collapse = ", "), "\n")
cat("MICE methods assigned:\n")
print(mice_imp$method)
cat("\nMICE sanity checks and slope summary:\n")
mice_imp1_slopes <- complete(mice_imp, 1)$edss_slope
mice_imp2_slopes <- complete(mice_imp, 2)$edss_slope
observed_idx     <- !is.na(features_continuous$edss_slope)

cat("\nMICE sanity checks:\n")
cat("  Observed slopes unchanged between imp 1 and 2:",
    all(mice_imp1_slopes[observed_idx] == mice_imp2_slopes[observed_idx]), "\n")
cat("  N patients with imputed slope:", sum(!observed_idx), "\n")
cat("  Imputed slope range imp 1:", round(range(mice_imp1_slopes[!observed_idx]), 3), "\n")
cat("  Imputed slope range imp 2:", round(range(mice_imp2_slopes[!observed_idx]), 3), "\n")
cat("  Imputed values differ between imp 1 and 2:",
    !all(mice_imp1_slopes[!observed_idx] == mice_imp2_slopes[!observed_idx]), "\n")
summary(complete(mice_imp, 1)$edss_slope)
```

## Strategy 4 - Bayesian multiple imputation (jomo)

```{r impute-jomo}
# jomo: Bayesian multiple imputation
# jomo package - Quartagno & Carpenter 2019

jomo_Y <- features_continuous %>%
  select(edss_slope) %>%
  as.data.frame()

jomo_X <- features_continuous %>%
  select(baseline_edss, edss_sd, visit_freq_yr,
         relapse_count, age_at_onset,
         GENDER, MSCOURSE_grouped,   
         cdp) %>%
  mutate(
    GENDER           = as.numeric(GENDER),
    MSCOURSE_grouped = as.numeric(MSCOURSE_grouped),
    cdp              = as.numeric(cdp)
  ) %>%
  as.data.frame()

set.seed(42)
jomo_imp <- jomo(
  Y        = jomo_Y,
  X        = jomo_X,
  nimp     = 20,
  nburn    = 1000,
  nbetween = 100
)

cat("jomo output columns:", paste(names(jomo_imp), collapse = ", "), "\n")
cat("jomo completed — 20 imputed datasets\n")

# --- sanity checks ---
jomo_imp1 <- jomo_imp %>% filter(Imputation == 1)
jomo_imp2 <- jomo_imp %>% filter(Imputation == 2)
observed_idx <- !is.na(features_continuous$edss_slope)
jomo_slopes1 <- as.numeric(jomo_imp1$edss_slope)
jomo_slopes2 <- as.numeric(jomo_imp2$edss_slope)

cat("\njomo sanity checks:\n")
cat("  Rows in imp 1:", nrow(jomo_imp1), "(should be", nrow(features_continuous), ")\n")
cat("  Rows in imp 2:", nrow(jomo_imp2), "(should be", nrow(features_continuous), ")\n")
cat("  Observed slopes identical between imp 1 and 2:",
    all(abs(jomo_slopes1[observed_idx] - jomo_slopes2[observed_idx]) < 1e-10), "\n")
cat("  N patients with imputed slope:", sum(!observed_idx), "\n")
cat("  Imputed slope range imp 1:", round(range(jomo_slopes1[!observed_idx]), 3), "\n")
cat("  Imputed slope range imp 2:", round(range(jomo_slopes2[!observed_idx]), 3), "\n")
cat("  Imputed values differ between imp 1 and 2:",
    !all(jomo_slopes1[!observed_idx] == jomo_slopes2[!observed_idx]), "\n")
```

## Compare imputed slope distributions

```{r impute-compare}
# plot slope distribution across all four strategies side by side
# showing imputation 1 of 20 as a representative example for MICE and jomo
# all 20 imputations produce similar distributions — showing all would overcrowd the plot
# key comparison: do imputed slopes preserve the shape and centre of the observed distribution?
slope_cc   <- features_cc   %>% mutate(strategy = "Complete Case")   %>% select(edss_slope, strategy)
slope_mean <- features_median %>% mutate(strategy = "Median Imputation") %>% select(edss_slope, strategy)

slope_mice <- complete(mice_imp, 1) %>%
  as_tibble() %>%
  mutate(strategy = "MICE") %>%
  select(edss_slope, strategy)

slope_jomo <- jomo_imp %>%
  filter(Imputation == 1) %>%
  as_tibble() %>%
  mutate(strategy = "Bayesian MI (jomo)",
         edss_slope = as.numeric(edss_slope)) %>%
  select(edss_slope, strategy)

slope_original <- features_continuous %>%
  filter(!is.na(edss_slope)) %>%
  mutate(strategy = "Original (observed)") %>%
  select(edss_slope, strategy)

all_slopes <- bind_rows(slope_original, slope_cc, slope_mean,
                         slope_mice, slope_jomo)

ggplot(all_slopes, aes(x = edss_slope, fill = strategy)) +
  geom_histogram(bins = 20, colour = "white", alpha = 0.8) +
  facet_wrap(~ strategy, ncol = 1) +
  geom_vline(xintercept = 0, colour = "#C85250", linetype = "dashed") +
  labs(
  title    = "EDSS Slope Distribution by Imputation Strategy",
  subtitle = paste0(
    "MICE and jomo: one representative imputed dataset shown (imputation 1 of 20).\n",
    "Complete Case: 52 patients with observed slopes only (22 excluded).\n",
    "Median Imputed: all 74 patients, missing slopes filled with median value.\n",
    "Original: the 52 observed slopes before any imputation.\n",
    "Dashed line at zero = no change in EDSS over the feature window."
  ),
  x = "EDSS slope (units per year)",
  y = "Count"
) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Ordinal matrix - no imputation needed

```{r ordinal-complete}
# ordinal matrix has zero missing values
# passes directly to modelling without imputation
cat("Ordinal matrix missing values:", sum(is.na(features_ordinal %>% select(-PATIENT_ID))), "\n")
cat("Ordinal matrix ready for modelling:", nrow(features_ordinal),
    "patients,", ncol(features_ordinal) - 2, "features\n")
```

## Save all imputed datasets

```{r save-imputed}
# save each strategy for use in modelling week
saveRDS(features_cc,   "features_cc.rds")
saveRDS(features_median, "features_median.rds") # median imputed
saveRDS(mice_imp,      "features_mice.rds")    # full mice object - pool results later
saveRDS(jomo_imp,      "features_jomo.rds")    # full jomo output

cat("All imputed datasets saved\n")
```
---

# Week 5 - Modelling Pipeline

> LASSO feature selection applied before each model.
> Three models: logistic regression, random forest, gradient boosting.
> Evaluation: AUC and Brier score with repeated cross-validation.
> Class imbalance handled via weighting (LR, GBM) and undersampling (RF).
> MICE and jomo predictions pooled by averaging predicted probabilities 
> across 20 imputations

## Load libraries for modelling

```{r modelling-libraries}
library(glmnet)
library(randomForest)
library(gbm)
library(pROC)
library(caret)
```

## Helper functions

```{r modelling-helpers}
# class weights computed from the full dataset passed in (not recomputed per fold)
# w_i = n / (2 * n_class) gives equal total weight to each outcome class
# when subsetted to a training fold, the weights retain the full-dataset class proportions
# (17 CDP, 57 non-CDP in the full 74-patient cohort)
# this is specified here and acknowledged as a methodological detail in the thesis
get_weights <- function(cdp) {
  n      <- length(cdp)
  n_pos  <- sum(cdp == 1)
  n_neg  <- sum(cdp == 0)
  w      <- ifelse(cdp == 1, n / (2 * n_pos), n / (2 * n_neg))
  return(w)
}

get_dummy_matrix <- function(df) {
  df %>%
    select(-PATIENT_ID, -cdp) %>%
    model.matrix(~ . - 1, data = .) %>%
    as.matrix()
}

evaluate_model <- function(actual, predicted_prob) {
  auc   <- as.numeric(auc(roc(actual, predicted_prob, quiet = TRUE)))
  brier <- mean((predicted_prob - actual)^2)
  list(auc = round(auc, 3), brier = round(brier, 3))
}
```

## LASSO feature selection

```{r lasso-cc}
# LASSO feature selection - run on full dataset before CV loop
# limitation: test fold patients' outcomes are visible during feature selection
# this introduces optimistic bias in reported AUC values (acknowledged in thesis)
# within-fold LASSO would be more correct but unstable with only 17 CDP events
set.seed(42)

run_lasso <- function(df, label, verbose = TRUE) {
  X   <- get_dummy_matrix(df)
  y   <- df$cdp
  w   <- get_weights(y)

  cv  <- cv.glmnet(X, y, family = "binomial", alpha = 1,
                   weights = w, nfolds = 5)

  selected <- coef(cv, s = "lambda.min")
  selected <- rownames(selected)[selected[, 1] != 0]
  selected <- selected[selected != "(Intercept)"]

  if (verbose) {
    cat(label, "- selected features:", paste(selected, collapse = ", "), "\n")
  }
  return(list(model = cv, features = selected, X = X, y = y, w = w))
}

lasso_cc     <- run_lasso(features_cc,     "Complete case")
lasso_median <- run_lasso(features_median, "Median imputed")
lasso_ord    <- run_lasso(features_ordinal,"Ordinal")
```

## Cross-validation pipeline

```{r cv-pipeline}
set.seed(42)

# separate folds for each dataset, sized to that dataset's own patient count
# (fewer patients in features_cc since complete-case dropped missing-slope patients)
folds_52  <- createFolds(features_cc$cdp,      k = 5, returnTrain = TRUE)
folds_74  <- createFolds(features_median$cdp,  k = 5, returnTrain = TRUE)
folds_ord <- createFolds(features_ordinal$cdp, k = 5, returnTrain = TRUE)

run_cv <- function(df, lasso_obj, label, folds) {

  X_all <- lasso_obj$X
  y_all <- lasso_obj$y
  w_all <- lasso_obj$w
  feats <- lasso_obj$features

  results_lr  <- c()
  results_rf  <- c()
  results_gbm <- c()
  brier_lr    <- c()
  brier_rf    <- c()
  brier_gbm   <- c()

  # storage for out-of-fold predictions, filled in as each fold completes
  probs_lr_oof  <- rep(NA, nrow(df))
  probs_rf_oof  <- rep(NA, nrow(df))
  probs_gbm_oof <- rep(NA, nrow(df))

  for (i in seq_along(folds)) {
    train_idx <- folds[[i]]
    test_idx  <- setdiff(seq_len(nrow(df)), train_idx)

    X_train <- X_all[train_idx, feats, drop = FALSE]
    X_test  <- X_all[test_idx,  feats, drop = FALSE]
    y_train <- y_all[train_idx]
    y_test  <- y_all[test_idx]
    # w_train: subset of weights computed from the full dataset in run_lasso()
    # not recomputed from the training fold — full-dataset class proportions are used
    w_train <- w_all[train_idx]

    train_df <- as.data.frame(X_train)
    train_df$cdp <- y_train
    test_df  <- as.data.frame(X_test)

    # logistic regression
    lr_fit  <- glm(cdp ~ ., data = train_df, family = binomial, weights = w_train)
    lr_pred <- predict(lr_fit, newdata = test_df, type = "response")
    ev_lr   <- evaluate_model(y_test, lr_pred)
    results_lr <- c(results_lr, ev_lr$auc)
    brier_lr   <- c(brier_lr,   ev_lr$brier)

    # random forest
    n_minority <- sum(y_train == 1)
    rf_fit <- randomForest(
      x        = X_train,
      y        = factor(y_train),
      ntree    = 500,
      sampsize = c(n_minority, n_minority)
    )
    rf_pred <- predict(rf_fit, newdata = X_test, type = "prob")[, 2]
    ev_rf   <- evaluate_model(y_test, rf_pred)
    results_rf <- c(results_rf, ev_rf$auc)
    brier_rf   <- c(brier_rf,   ev_rf$brier)

    # gradient boosting
    gbm_fit  <- gbm(
      cdp ~ .,
      data              = train_df,
      distribution      = "bernoulli",
      n.trees           = 500,
      interaction.depth = 2,
      shrinkage         = 0.01,
      n.minobsinnode    = 1,
      bag.fraction      = 0.5,
      weights           = w_train,
      verbose           = FALSE
    )
    gbm_pred <- predict(gbm_fit, newdata = test_df, n.trees = 500, type = "response")
    ev_gbm   <- evaluate_model(y_test, gbm_pred)
    results_gbm <- c(results_gbm, ev_gbm$auc)
    brier_gbm   <- c(brier_gbm,   ev_gbm$brier)

    # save this fold's test-set predictions into the full-length vector
    probs_lr_oof[test_idx]  <- as.numeric(lr_pred)
    probs_rf_oof[test_idx]  <- as.numeric(rf_pred)
    probs_gbm_oof[test_idx] <- as.numeric(gbm_pred)
  }

  cat("\n===", label, "===\n")
  cat("LR  AUC:", round(mean(results_lr),  3), "SD:", round(sd(results_lr),  3), "| Brier:", round(mean(brier_lr), 3), "\n")
  cat("RF  AUC:", round(mean(results_rf),  3), "SD:", round(sd(results_rf),  3), "| Brier:", round(mean(brier_rf), 3), "\n")
  cat("GBM AUC:", round(mean(results_gbm), 3), "SD:", round(sd(results_gbm), 3), "| Brier:", round(mean(brier_gbm), 3), "\n")

  # return the summary table plus the OOF predictions for calibration plots
  list(
    table = tibble(
      strategy   = label,
      model      = c("Logistic Regression", "Random Forest", "Gradient Boosting"),
      mean_auc   = c(mean(results_lr),  mean(results_rf),  mean(results_gbm)),
      sd_auc     = c(sd(results_lr),    sd(results_rf),    sd(results_gbm)),
      mean_brier = c(mean(brier_lr),    mean(brier_rf),    mean(brier_gbm))
    ),
    probs_lr  = probs_lr_oof,
    probs_rf  = probs_rf_oof,
    probs_gbm = probs_gbm_oof
  )
}

results_cc     <- run_cv(features_cc,      lasso_cc,     "Complete Case - Continuous",  folds_52)
results_median <- run_cv(features_median,  lasso_median, "Median Imputed - Continuous", folds_74)
results_ord    <- run_cv(features_ordinal, lasso_ord,    "No Imputation - Ordinal",     folds_ord)
```

## MICE and jomo - pooled modelling

```{r mice-jomo-pooling}
set.seed(42)

# pooling probabilities across M imputations
# run 5-fold CV on each imputed dataset, average the predicted probs per
# patient across the M datasets, then compute AUC/Brier from the averaged
# probabilities. SD across the M per-imputation AUCs gives the uncertainty.

run_cv_pooled <- function(imp_object, imp_type, label, M = 20) {

  all_probs_lr  <- matrix(NA, nrow = nrow(features_continuous), ncol = M)
  all_probs_rf  <- matrix(NA, nrow = nrow(features_continuous), ncol = M)
  all_probs_gbm <- matrix(NA, nrow = nrow(features_continuous), ncol = M)

  auc_lr_each  <- c()
  auc_rf_each  <- c()
  auc_gbm_each <- c()

  for (m in 1:M) {

    # extract the mth complete dataset
    if (imp_type == "mice") {
      df_m <- complete(imp_object, m) %>%
        as_tibble() %>%
        mutate(PATIENT_ID = features_continuous$PATIENT_ID)
    } else {
      df_m <- imp_object %>%
        filter(Imputation == m) %>%
        select(edss_slope, baseline_edss, edss_sd, visit_freq_yr,
               relapse_count, age_at_onset, cdp) %>%
        as_tibble() %>%
        mutate(
          across(everything(), as.numeric),
          GENDER           = features_continuous$GENDER,
          MSCOURSE_grouped = features_continuous$MSCOURSE_grouped,
          PATIENT_ID       = features_continuous$PATIENT_ID
        )
    }

    if (m == 1 && imp_type == "jomo") {
      cat("Row order check for jomo - should say TRUE for all", nrow(df_m), "rows:\n")
      print(nrow(df_m) == nrow(features_continuous))
    }
    # LASSO run on each full imputed dataset (all 74 patients) before CV folds
    # same limitation applies as for the non-pooled strategies:
    # feature selection sees all outcomes including those in subsequent test folds
    lasso_m <- run_lasso(df_m, paste(label, "imp", m), verbose = FALSE)

    X_all <- lasso_m$X
    y_all <- lasso_m$y
    w_all <- lasso_m$w
    feats <- lasso_m$features

    probs_lr_m  <- rep(NA, nrow(features_continuous))
    probs_rf_m  <- rep(NA, nrow(features_continuous))
    probs_gbm_m <- rep(NA, nrow(features_continuous))
    
    for (i in seq_along(folds_74)) {
      train_idx <- folds_74[[i]]
      test_idx  <- setdiff(seq_len(nrow(features_continuous)), train_idx)

      X_train <- X_all[train_idx, feats, drop = FALSE]
      X_test  <- X_all[test_idx,  feats, drop = FALSE]
      y_train <- y_all[train_idx]
      y_test  <- y_all[test_idx]
      w_train <- w_all[train_idx]

      train_df <- as.data.frame(X_train)
      train_df$cdp <- y_train
      test_df  <- as.data.frame(X_test)

      # logistic regression
      lr_fit <- glm(cdp ~ ., data = train_df,
                    family = binomial, weights = w_train)
      probs_lr_m[test_idx] <- predict(lr_fit, newdata = test_df,
                                       type = "response")

      # random forest
      n_min <- sum(y_train == 1)
      rf_fit <- randomForest(
        x        = X_train,
        y        = factor(y_train),
        ntree    = 500,
        sampsize = c(n_min, n_min)
      )
      probs_rf_m[test_idx] <- predict(rf_fit, newdata = X_test,
                                       type = "prob")[, 2]

      # gradient boosting
      gbm_fit <- gbm(
        cdp ~ .,
        data              = train_df,
        distribution      = "bernoulli",
        n.trees           = 500,
        interaction.depth = 2,
        shrinkage         = 0.01,
        n.minobsinnode    = 1,
        bag.fraction      = 0.5,
        weights           = w_train,
        verbose           = FALSE
      )
      probs_gbm_m[test_idx] <- predict(gbm_fit, newdata = test_df,
                                        n.trees = 500, type = "response")
    }

    all_probs_lr[,  m] <- probs_lr_m
    all_probs_rf[,  m] <- probs_rf_m
    all_probs_gbm[, m] <- probs_gbm_m

    y <- y_all
    auc_lr_each  <- c(auc_lr_each,
                      as.numeric(auc(roc(y, probs_lr_m,  quiet = TRUE))))
    auc_rf_each  <- c(auc_rf_each,
                      as.numeric(auc(roc(y, probs_rf_m,  quiet = TRUE))))
    auc_gbm_each <- c(auc_gbm_each,
                      as.numeric(auc(roc(y, probs_gbm_m, quiet = TRUE))))
  }

  # pool: average predicted probabilities across M imputations
  pooled_lr  <- rowMeans(all_probs_lr,  na.rm = TRUE)
  pooled_rf  <- rowMeans(all_probs_rf,  na.rm = TRUE)
  pooled_gbm <- rowMeans(all_probs_gbm, na.rm = TRUE)

  y <- features_continuous$cdp

  ev_lr  <- evaluate_model(y, pooled_lr)
  ev_rf  <- evaluate_model(y, pooled_rf)
  ev_gbm <- evaluate_model(y, pooled_gbm)

  cat("\n===", label, "===\n")
  cat("LR  pooled AUC:", ev_lr$auc,
      "| SD across imputations:", round(sd(auc_lr_each), 3),
      "| Brier:", ev_lr$brier, "\n")
  cat("RF  pooled AUC:", ev_rf$auc,
      "| SD across imputations:", round(sd(auc_rf_each), 3),
      "| Brier:", ev_rf$brier, "\n")
  cat("GBM pooled AUC:", ev_gbm$auc,
      "| SD across imputations:", round(sd(auc_gbm_each), 3),
      "| Brier:", ev_gbm$brier, "\n")
  # note on SD similarity across models for MICE and jomo:
  # the SD here measures variability across the 20 imputed datasets, NOT across CV folds
  # all three models see the same 20 versions of the slope feature
  # so variation in AUC across imputations is driven by the imputed slope values, not the model
  # this means the SD will be similar across LR, RF, and GBM for the same imputation strategy
  # contrast with complete case and median imputed: those SDs are across CV folds (fold-level noise)
  # which is why complete case GBM SD is 0.248 (fold instability) while jomo GBM SD is 0.030

  results_table <- tibble(
    strategy   = label,
    model      = c("Logistic Regression", "Random Forest", "Gradient Boosting"),
    pooled_auc = c(ev_lr$auc,   ev_rf$auc,   ev_gbm$auc),
    sd_auc     = c(round(sd(auc_lr_each), 3),
                   round(sd(auc_rf_each), 3),
                   round(sd(auc_gbm_each), 3)),
    brier      = c(ev_lr$brier, ev_rf$brier, ev_gbm$brier)
  )

  return(list(table = results_table,
              pooled_lr = pooled_lr, pooled_rf = pooled_rf, pooled_gbm = pooled_gbm))
}

mice_out <- run_cv_pooled(mice_imp, "mice", "MICE - Continuous")
jomo_out <- run_cv_pooled(jomo_imp, "jomo", "Bayesian MI (jomo) - Continuous")

results_mice <- mice_out$table
results_jomo <- jomo_out$table
```

## Complete comparison matrix

```{r comparison-matrix}
all_results_final <- bind_rows(
  results_cc$table %>% rename(pooled_auc = mean_auc, brier = mean_brier),
  results_median$table %>% rename(pooled_auc = mean_auc, brier = mean_brier),
  results_ord$table    %>% rename(pooled_auc = mean_auc, brier = mean_brier),
  results_mice,
  results_jomo
) %>%
  mutate(
    pooled_auc = round(pooled_auc, 3),
    sd_auc     = round(sd_auc,     3),
    brier      = round(brier,      3)
  )
print(as.data.frame(all_results_final))
all_results_final %>%
  kable(col.names = c("Strategy", "Model", "AUC", "SD", "Brier"),
        caption   = "Full comparison matrix - all imputation strategies and models") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(all_results_final$strategy == "No Imputation - Ordinal"),
           background = "#EBF5E8")
```

## Final results plot

```{r final-plot}
ggplot(all_results_final, aes(x = model, y = pooled_auc, fill = strategy)) +
  geom_col(position = "dodge", colour = "white", alpha = 0.85) +
  coord_cartesian(ylim = c(0.5, 1.0)) +
  geom_errorbar(aes(ymin = pooled_auc - sd_auc,
                    ymax = pooled_auc + sd_auc),
                position = position_dodge(0.9), width = 0.25) +
  scale_fill_manual(values = c("#2E4D7B", "#C85250", "#5A8A5A",
                               "#E8A020", "#7B4F9E")) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  labs(title    = "Full Comparison Matrix - AUC by Strategy and Model",
       subtitle = "Error bars = SD across imputations or CV folds. Dashed = random chance.",
       x = "", y = "AUC", fill = "Strategy") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 20, hjust = 1))
```

## Calibration Plots

```{r calibration-plots}
library(ggplot2)

calibration_plot <- function(probs, actual, title) {
  df <- tibble(prob = probs, actual = actual) %>%
    mutate(bin = cut(prob, breaks = seq(0, 1, by = 0.2),
                     include.lowest = TRUE)) %>%
    group_by(bin) %>%
    summarise(
      mean_pred = mean(prob),
      mean_obs  = mean(actual),
      n         = n()
    )
  print(df)
  ggplot(df, aes(x = mean_pred, y = mean_obs)) +
    geom_point(aes(size = n), colour = "#2E4D7B", alpha = 0.8) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "#C85250") +
    scale_x_continuous(limits = c(0, 1)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(title    = title,
         subtitle = "Dashed line = perfect calibration",
         x = "Mean predicted probability",
         y = "Observed event rate",
         size = "N patients") +
    theme_minimal(base_size = 12)
}
calibration_plot(mice_out$pooled_rf, features_continuous$cdp,
                  "Calibration - Random Forest (MICE pooled)")

calibration_plot(jomo_out$pooled_rf, features_continuous$cdp,
                  "Calibration - Random Forest (jomo pooled)")
calibration_plot(results_ord$probs_rf, features_ordinal$cdp,
                  "Calibration - Random Forest (Ordinal)")
```

## Feature Importance

```{r feature-importance}
# train final RF on full ordinal dataset (best performing arm)
# use all data not CV splits for importance plot

X_full <- get_dummy_matrix(features_ordinal)
y_full <- features_ordinal$cdp
feats  <- lasso_ord$features

rf_final <- randomForest(
  x        = X_full[, feats, drop = FALSE],
  y        = factor(y_full),
  ntree    = 500,
  importance = TRUE
)

importance_df <- importance(rf_final) %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  arrange(desc(MeanDecreaseGini))
print(importance_df)
importance_df %>%
  kable(caption = "Feature importance - ordinal RF (full dataset)",
        digits  = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

ggplot(importance_df, aes(x = reorder(feature, MeanDecreaseGini),
                           y = MeanDecreaseGini)) +
  geom_col(fill = "#2E4D7B", alpha = 0.85) +
  coord_flip() +
  labs(title = "Feature Importance - Ordinal Random Forest",
       x = "", y = "Mean Decrease in Gini") +
  theme_minimal(base_size = 12)
```
## GBM Feature Importance

```{r gbm-importance}
gbm_final <- gbm(
  cdp ~ .,
  data = as.data.frame(cbind(
    X_full[, feats, drop = FALSE],
    cdp = y_full
  )),
  distribution      = "bernoulli",
  n.trees           = 500,
  interaction.depth = 2,
  shrinkage         = 0.01,
  n.minobsinnode    = 1,
  bag.fraction      = 0.5,
  verbose           = FALSE
)

gbm_imp <- summary(gbm_final, plotit = FALSE) %>%
  as_tibble() %>%
  rename(feature = var, importance = rel.inf) %>%
  arrange(desc(importance))
print(gbm_imp)
gbm_imp %>%
  kable(caption = "Feature importance - ordinal GBM (full dataset)",
        digits  = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

ggplot(gbm_imp, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#C85250", alpha = 0.85) +
  coord_flip() +
  labs(title = "Feature Importance - Ordinal GBM",
       x = "", y = "Relative Influence (%)") +
  theme_minimal(base_size = 12)
```

## Clinical Translation - Youden Threshold

```{r clinical-transition}
library(pROC)

roc_ord_rf <- roc(features_ordinal$cdp, results_ord$probs_rf, quiet = TRUE)
best_ord <- coords(roc_ord_rf, "best", best.method = "youden",
                    ret = c("threshold", "sensitivity", "specificity",
                            "ppv", "npv", "tp", "fp", "tn", "fn"))
print(best_ord)

roc_mice_gbm <- roc(features_continuous$cdp, mice_out$pooled_gbm, quiet = TRUE)
best_mice <- coords(roc_mice_gbm, "best", best.method = "youden",
                     ret = c("threshold", "sensitivity", "specificity",
                             "ppv", "npv", "tp", "fp", "tn", "fn"))
print(best_mice)
```

## Continuous Arm

```{r continuous-arm}
mice_imp1 <- complete(mice_imp, 1) %>%
  as_tibble() %>%
  mutate(PATIENT_ID = features_continuous$PATIENT_ID)

lasso_mice1 <- run_lasso(mice_imp1, "MICE imp 1", verbose = FALSE)
X_mice1 <- lasso_mice1$X[, lasso_mice1$features, drop = FALSE]
y_mice1 <- lasso_mice1$y

gbm_mice_final <- gbm(
  cdp ~ .,
  data = as.data.frame(cbind(X_mice1, cdp = y_mice1)),
  distribution      = "bernoulli",
  n.trees           = 500,
  interaction.depth = 2,
  shrinkage         = 0.01,
  n.minobsinnode    = 1,
  bag.fraction      = 0.5,
  verbose           = FALSE
)

gbm_mice_imp <- summary(gbm_mice_final, plotit = FALSE) %>%
  as_tibble() %>%
  rename(feature = var, importance = rel.inf) %>%
  arrange(desc(importance))
print(gbm_mice_imp)

ggplot(gbm_mice_imp, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#5A8A5A", alpha = 0.85) +
  coord_flip() +
  labs(title = "Feature Importance - MICE GBM (Continuous, imputation 1)",
       x = "", y = "Relative Influence (%)") +
  theme_minimal(base_size = 12)

```
## Calibration - Worst Performing Combination

```{r calibration-worst}
# worst performing: complete case GBM
# uses out-of-fold predicted probabilities from run_cv, not in-sample predictions
# out-of-fold predictions come from 5-fold CV on the 52-patient complete case cohort
calibration_plot(
  probs  = results_cc$probs_gbm,
  actual = features_cc$cdp,
  title  = "Calibration - GBM (Complete Case) - Worst Performing"
)
```

## MNAR Sensitivity

```{r mnar-sensitivity}
# tipping point analysis for MNAR assumption
# shift the imputed slopes by a delta and see how AUC changes
# if conclusions hold up to delta = 0.5 EDSS units per year, MNAR is not a concern
missing_patients <- features_continuous$PATIENT_ID[is.na(features_continuous$edss_slope)]
deltas <- seq(-0.5, 0.5, by = 0.1)
auc_sensitivity <- c()

for (delta in deltas) {
  df_shifted <- features_median %>%
    mutate(edss_slope = if_else(
      PATIENT_ID %in% missing_patients,
      edss_slope + delta,
      edss_slope
    ))

  lasso_s <- run_lasso(df_shifted, paste("delta", delta))
  X_s <- lasso_s$X
  y_s <- lasso_s$y

  lr_fit  <- glm(cdp ~ .,
                 data    = as.data.frame(cbind(X_s[, lasso_s$features,
                                                    drop = FALSE],
                                               cdp = y_s)),
                 family  = binomial,
                 weights = lasso_s$w)

  preds <- predict(lr_fit, type = "response")
  auc_sensitivity <- c(auc_sensitivity,
                       as.numeric(auc(roc(y_s, preds, quiet = TRUE))))
}
print(tibble(delta = deltas, auc = round(auc_sensitivity, 3)))
tibble(delta = deltas, auc = round(auc_sensitivity, 3)) %>%
  kable(col.names = c("Slope shift (EDSS/yr)", "In-sample LR AUC"),
        caption   = "MNAR sensitivity - effect of shifting imputed slopes (in-sample AUC, not directly comparable to CV results above)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```
---

## Export Figures for Thesis

```{r export-figures, include=FALSE}
figures_path <- "Figures"
dir.create(figures_path, showWarnings = FALSE)


## EDA figures


# Patient demographics grid (thesis: fig:patientdemographics)
p_dem1 <- ggplot(patients_key, aes(x = age_at_onset)) +
  geom_histogram(bins = 30, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  labs(title = "Age at Symptom Onset", x = "Age (years)", y = "Number of patients") +
  theme_minimal(base_size = 13)

p_dem2 <- ggplot(patients_key %>% filter(!is.na(GENDER)),
                 aes(x = GENDER, fill = GENDER)) +
  geom_bar(colour = "white", alpha = 0.85) +
  scale_fill_manual(values = c("#2E4D7B", "#C85250", "#5A8A5A")) +
  labs(title = "Gender Distribution", x = "", y = "Number of patients") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p_dem3 <- ggplot(patients_key %>% filter(!is.na(MSCOURSE_1)),
                 aes(x = fct_infreq(MSCOURSE_1), fill = MSCOURSE_1)) +
  geom_bar(colour = "white", alpha = 0.85) +
  scale_fill_viridis_d() +
  labs(title = "Initial Disease Course", x = "", y = "Number of patients") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        axis.text.x = element_text(hjust = 0.5))

p_dem4 <- ggplot(patients_key %>% filter(!is.na(diagnostic_delay_yrs),
                                          diagnostic_delay_yrs >= 0,
                                          diagnostic_delay_yrs < 30),
                 aes(x = diagnostic_delay_yrs)) +
  geom_histogram(bins = 25, fill = "#5A8A5A", colour = "white", alpha = 0.85) +
  labs(title = "Diagnostic Delay", x = "Years", y = "Number of patients") +
  theme_minimal(base_size = 13)

p_demographics <- ggarrange(p_dem1, p_dem2, p_dem3, p_dem4, ncol = 2, nrow = 2)
ggsave(file.path(figures_path, "patient_demographics.png"),
       p_demographics, width = 12, height = 9, dpi = 300)

# Visits per patient (thesis: fig:visitsperpatient)
p_visits <- ggplot(visits_per_patient, aes(x = n_visits)) +
  geom_histogram(bins = 30, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  geom_vline(aes(xintercept = median(n_visits)),
             colour = "#C85250", linetype = "dashed", linewidth = 1) +
  annotate("text",
           x     = median(visits_per_patient$n_visits) + 1,
           y     = Inf, vjust = 2,
           label = paste("Median =", median(visits_per_patient$n_visits)),
           colour = "#C85250", size = 4) +
  labs(title = "Distribution of Number of Visits per Patient",
       x = "Number of visits", y = "Number of patients") +
  theme_minimal(base_size = 13)
ggsave(file.path(figures_path, "visits_per_patient.png"),
       p_visits, width = 10, height = 6, dpi = 300)

# Inter-visit gap (thesis: fig:visitspacing)
p_gaps <- ggplot(visit_gaps %>% filter(gap_days < 730), aes(x = gap_months)) +
  geom_histogram(bins = 40, fill = "#5A8A5A", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 6,  colour = "#C85250", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 12, colour = "#2E4D7B", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 6.5,  y = Inf, vjust = 2,
           label = "6 months",  colour = "#C85250", size = 3.5) +
  annotate("text", x = 12.5, y = Inf, vjust = 2,
           label = "12 months", colour = "#2E4D7B", size = 3.5) +
  labs(title    = "Distribution of Inter-Visit Gaps (months)",
       subtitle = "Confirms irregular visit spacing",
       x = "Gap between consecutive visits (months)", y = "Count") +
  theme_minimal(base_size = 13)
ggsave(file.path(figures_path, "visit_spacing.png"),
       p_gaps, width = 10, height = 6, dpi = 300)

# EDSS missingness over time (thesis: fig:edssmissingtime)
p_edss_miss <- ggplot(edss_missing_time,
                      aes(x = DURATION_OF_MS_AT_VISIT_ROUNDED, y = pct_missing)) +
  geom_col(fill = "#C85250", colour = "white", alpha = 0.85) +
  geom_hline(yintercept = 45, linetype = "dashed", colour = "#2E4D7B", linewidth = 1) +
  annotate("text", x = 15, y = 47, label = "Overall 45% missing",
           colour = "#2E4D7B", size = 3.5) +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +
  labs(title    = "EDSS Missingness by Year of Disease Duration",
       subtitle = "If missingness increases over time, this supports MNAR",
       x = "Years since symptom onset", y = "% of visits missing EDSS") +
  theme_minimal(base_size = 13)
ggsave(file.path(figures_path, "edss_missing_time.png"),
       p_edss_miss, width = 10, height = 6, dpi = 300)

# Missing data overview (thesis: fig:missingoverview)
p_missing <- gg_miss_var(
  analysis_vars %>% select(-PATIENT_ID), show_pct = TRUE) +
  labs(title    = "Missing Data - Combined Key Variables",
       subtitle = "Percentage of values missing per variable") +
  theme_minimal(base_size = 12)
ggsave(file.path(figures_path, "missing_overview.png"),
       p_missing, width = 10, height = 8, dpi = 300)

# EDSS trajectory (thesis: fig:edsstrajectory)
p_edss_traj <- ggplot(edss_trajectory,
                      aes(x = DURATION_OF_MS_AT_VISIT_ROUNDED)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#2E4D7B", alpha = 0.2) +
  geom_line(aes(y = median_edss), colour = "#2E4D7B", linewidth = 1.2) +
  geom_point(aes(y = median_edss, size = n_visits), colour = "#2E4D7B") +
  scale_size_continuous(range = c(2, 6), name = "N visits") +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), limits = c(0, 10)) +
  labs(title    = "Median EDSS by Years Since Symptom Onset",
       subtitle = "Shaded band = IQR (25th to 75th percentile)",
       x = "Years since symptom onset", y = "EDSS") +
  theme_minimal(base_size = 13)
ggsave(file.path(figures_path, "edss_trajectory.png"),
       p_edss_traj, width = 10, height = 6, dpi = 300)

# Relapse plots (thesis: fig:relapseplots)
p_rel1 <- ggplot(relapses_per_patient, aes(x = n_relapses)) +
  geom_histogram(bins = 20, fill = "#C85250", colour = "white", alpha = 0.85) +
  labs(title = "Relapses per Patient",
       x = "Number of relapses", y = "Number of patients") +
  theme_minimal(base_size = 13)

p_rel2 <- ggplot(relapses_with_time, aes(x = time_since_onset_yrs)) +
  geom_histogram(bins = 30, fill = "#C85250", colour = "white", alpha = 0.85) +
  labs(title = "Relapse Timing Since Symptom Onset",
       x = "Years since onset", y = "Number of relapses") +
  theme_minimal(base_size = 13)

ggsave(file.path(figures_path, "relapse_plots.png"),
       ggarrange(p_rel1, p_rel2, ncol = 2),
       width = 12, height = 6, dpi = 300)

# Follow-up length (thesis: fig:followuplength)
p_followup <- ggplot(followup %>% filter(!is.na(followup_from_onset_yrs)),
                     aes(x = followup_from_onset_yrs)) +
  geom_histogram(bins = 30, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 2, colour = "#C85250", linetype = "dashed", linewidth = 1.2) +
  geom_vline(xintercept = 5, colour = "#5A8A5A", linetype = "dashed", linewidth = 1.2) +
  geom_vline(xintercept = 7, colour = "#E8A020", linetype = "dashed", linewidth = 1.2) +
  annotate("text", x = 2.2, y = Inf, vjust = 2, label = "2 yrs",
           colour = "#C85250", size = 3.5) +
  annotate("text", x = 5.2, y = Inf, vjust = 2, label = "5 yrs",
           colour = "#5A8A5A", size = 3.5) +
  annotate("text", x = 7.2, y = Inf, vjust = 2, label = "7 yrs",
           colour = "#E8A020", size = 3.5) +
  labs(title    = "Distribution of Patient Follow-Up Length (from symptom onset)",
       subtitle = "Dashed lines show proposed feature and prediction windows",
       x = "Years from symptom onset to last visit", y = "Number of patients") +
  theme_minimal(base_size = 13)
ggsave(file.path(figures_path, "followup_length.png"),
       p_followup, width = 10, height = 6, dpi = 300)


## Feature engineering figures


# Feature distributions grid (thesis: fig:featuredistributions)
p_f1 <- ggplot(features_continuous %>% filter(!is.na(baseline_edss)),
               aes(x = baseline_edss)) +
  geom_histogram(bins = 15, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  labs(title = "Baseline EDSS", x = "EDSS", y = "Patients") +
  theme_minimal(base_size = 12)

p_f2 <- ggplot(features_continuous %>% filter(!is.na(edss_slope)),
               aes(x = edss_slope)) +
  geom_histogram(bins = 15, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, colour = "#C85250", linetype = "dashed") +
  labs(title = "EDSS Slope", x = "EDSS units per year", y = "Patients") +
  theme_minimal(base_size = 12)

p_f3 <- ggplot(features_continuous %>% filter(!is.na(edss_sd)),
               aes(x = edss_sd)) +
  geom_histogram(bins = 15, fill = "#2E4D7B", colour = "white", alpha = 0.85) +
  labs(title = "EDSS Variability (SD)", x = "SD", y = "Patients") +
  theme_minimal(base_size = 12)

p_f4 <- ggplot(features_continuous %>% filter(!is.na(age_at_onset)),
               aes(x = age_at_onset)) +
  geom_histogram(bins = 20, fill = "#5A8A5A", colour = "white", alpha = 0.85) +
  labs(title = "Age at Onset", x = "Years", y = "Patients") +
  theme_minimal(base_size = 12)

p_f5 <- ggplot(features_continuous, aes(x = relapse_count)) +
  geom_histogram(bins = 10, fill = "#C85250", colour = "white", alpha = 0.85) +
  labs(title = "Relapse Count in Feature Window",
       x = "Number of relapses", y = "Patients") +
  theme_minimal(base_size = 12)

p_f6 <- ggplot(features_continuous %>% filter(!is.na(MSCOURSE_grouped)),
               aes(x = MSCOURSE_grouped, fill = MSCOURSE_grouped)) +
  geom_bar(colour = "white", alpha = 0.85) +
  scale_fill_manual(values = c("#2E4D7B", "#C85250")) +
  labs(title = "MS Course", x = "", y = "Patients") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

ggsave(file.path(figures_path, "feature_distributions.png"),
       ggarrange(p_f1, p_f2, p_f3, p_f4, p_f5, p_f6, ncol = 2, nrow = 3),
       width = 12, height = 14, dpi = 300)

# Slope comparison by imputation strategy (thesis: fig:imputecompare)
p_slope_compare <- ggplot(all_slopes, aes(x = edss_slope, fill = strategy)) +
  geom_histogram(bins = 20, colour = "white", alpha = 0.8) +
  facet_wrap(~ strategy, ncol = 1) +
  geom_vline(xintercept = 0, colour = "#C85250", linetype = "dashed") +
  labs(  title = "EDSS Slope Distribution by Imputation Strategy",
       x = "EDSS slope (units per year)", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
ggsave(file.path(figures_path, "slope_comparison.png"),
       p_slope_compare, width = 8, height = 14, dpi = 300)


## Modelling figures


# Full AUC comparison matrix (thesis: fig:finalauc)
p_auc_final <- ggplot(all_results_final,
                      aes(x = model, y = pooled_auc, fill = strategy)) +
  geom_col(position = "dodge", colour = "white", alpha = 0.85) +
  coord_cartesian(ylim = c(0.5, 1.0)) +
  geom_errorbar(aes(ymin = pooled_auc - sd_auc,
                    ymax = pooled_auc + sd_auc),
                position = position_dodge(0.9), width = 0.25) +
  scale_fill_manual(values = c("#2E4D7B", "#C85250", "#5A8A5A",
                               "#E8A020", "#7B4F9E")) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  labs(title    = "Full Comparison Matrix - AUC by Strategy and Model",
       subtitle = "Error bars = SD across imputations or CV folds. Dashed = random chance.",
       x = "", y = "AUC", fill = "Strategy") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 20, hjust = 1))
ggsave(file.path(figures_path, "final_auc_comparison.png"),
       p_auc_final, width = 12, height = 7, dpi = 300)

# Calibration MICE RF (thesis: fig:calibmicerf)
ggsave(file.path(figures_path, "calibration_mice_rf.png"),
       calibration_plot(mice_out$pooled_rf, features_continuous$cdp,
                        "Calibration - Random Forest (MICE pooled)"),
       width = 7, height = 6, dpi = 300)

# Calibration jomo RF (thesis: fig:calibjomorf)
ggsave(file.path(figures_path, "calibration_jomo_rf.png"),
       calibration_plot(jomo_out$pooled_rf, features_continuous$cdp,
                        "Calibration - Random Forest (jomo pooled)"),
       width = 7, height = 6, dpi = 300)
ggsave(file.path(figures_path, "calibration_ordinal_rf.png"),
       calibration_plot(results_ord$probs_rf, features_ordinal$cdp,
                         "Calibration - Random Forest (Ordinal)"),
       width = 6, height = 5, dpi = 300)

# Calibration worst (thesis: fig:calibworst)
ggsave(
  filename = file.path(figures_path, "calibration_cc_gbm.png"),
  plot     = calibration_plot(
               probs  = results_cc$probs_gbm,
               actual = features_cc$cdp,
               title  = "Calibration - GBM (Complete Case) - Worst Performing"
             ),
  width    = 7, 
  height   = 6, 
  dpi      = 300
)

# RF feature importance (thesis: fig:rfimportance)
p_rf_imp <- ggplot(importance_df,
                   aes(x = reorder(feature, MeanDecreaseGini),
                       y = MeanDecreaseGini)) +
  geom_col(fill = "#2E4D7B", alpha = 0.85) +
  coord_flip() +
  labs(title = "Feature Importance - Ordinal Random Forest",
       x = "", y = "Mean Decrease in Gini") +
  theme_minimal(base_size = 12)
ggsave(file.path(figures_path, "rf_feature_importance.png"),
       p_rf_imp, width = 8, height = 5, dpi = 300)

# GBM feature importance (thesis: fig:gbmimportance)
p_gbm_imp <- ggplot(gbm_imp,
                    aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#C85250", alpha = 0.85) +
  coord_flip() +
  labs(title = "Feature Importance - Ordinal GBM",
       x = "", y = "Relative Influence (%)") +
  theme_minimal(base_size = 12)
ggsave(file.path(figures_path, "gbm_feature_importance.png"),
       p_gbm_imp, width = 8, height = 5, dpi = 300)

# MICE GBM feature importance, continuous arm (thesis: fig:gbmmiceimportance)
p_gbm_mice_imp <- ggplot(gbm_mice_imp,
                         aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#5A8A5A", alpha = 0.85) +
  coord_flip() +
  labs(title = "Feature Importance - MICE GBM (Continuous)",
       x = "", y = "Relative Influence (%)") +
  theme_minimal(base_size = 12)
ggsave(file.path(figures_path, "gbm_mice_feature_importance.png"),
       p_gbm_mice_imp, width = 9, height = 5, dpi = 300)


## Confirm all files saved

saved <- list.files(figures_path, pattern = "\\.png$")
cat("Figures saved to Figures/ folder:\n")
cat(paste(" -", saved), sep = "\n")
cat("\nTotal:", length(saved), "files\n")
```

---

*Generated: `r format(Sys.Date(), "%d %B %Y")` | Student 25044419 | Weeks 1–5*
